Hierarchical Reverberation Mapping

### 2013-11-01 Paper/ms.tex
Slight wording change
Various edits

% Bold symbols
\renewcommand{\btheta}{\boldsymbol{\theta}}
\newcommand{\bx}{\boldsymbol{x}}

@@

\section{Introduction}
Reverberation mapping (RM) is an important technique for the study
of active galactic nuclei (AGN). The technique is based on the temporal
fluctuations of the central continuum source, and the subsequent response
of the broad line region (BLR) emission. The time delay between the continuum
and the broad line fluctuations provides an estimate of the size of the BLR,
and can also be used to estimate the black hole mass \citep{peterson}.

RM is a observationally intensive, requiring observations of an AGNN
over a period of
a few tens of days \citep{2013ApJ...769..128B}. As a result, many authors have
studied the data analysis techniques involved in RM, and substantial
advances have been made in recent years.
The data analysis methods introduced range from those that attempt to infer the
transfer function \citep[the distribution of lags in a single object,
e.g.][]{1995ApJ...440..166K, 2011ApJ...735...80Z},
the velocity-resolved transfer function
\citep{2010ApJ...720L..46B}, or the physical structure of the BLR itself
\citep{pancoast, arp151, pancoast2, 2013arXiv1310.3907L}.

There are many subtleties and challenges involved in reverberation mapping,
that we will ignore
for the purposes of this letter. We note two of them here for completeness. Firstly,
the mean lag $\bar{\tau} = \int \tau\Psi(\tau)\, d\tau$, where $\Psi(\tau)$
is the normalised transfer function, is not equal to $c$ times the mean radius of the BLR
distribution, $\bar{r} = \int \sqrt{x^2+y^2+z^2}\rho(x, y, z)\, d^3 \mathbf{x}$.
Secondly, the mean lag $\bar{\tau}$ is not equivalent to the peak of the
cross-correlation function, except in the case of very narrow
transfer functions. Direct physical modelling of the BLR resolves these
issues \citep{pancoast, arp151, 2013arXiv1310.3907L}.

@@

In this letter, we revisit the idea of ``stacked'' reverberation mapping (i.e.
using a sample of AGN with sparse data to infer something about the sample)
from a Bayesian perspective. We develop a Bayesian hierarchical model
\citep{loredo, kelly, extreme_deconvolution, 2012AJ....143...90S, 2012arXiv1208.3036L, 2013arXiv1310.5177B, 2013AJ....146....7B}.
which allows us to measure

@@

\section{Combining Inferences About Multiple Objects}\label{sec:multiple}
Consider a sample of $N$ objects, each of which has parameters $\theta_i$.
If the $i$th object is analysed on its own, the inference about 
its parameters $\theta_i$ is described by a posterior distribution
\begin{eqnarray}
p(\theta_i | x_i) \propto \pi(\theta_i)p(x_i | \theta_i)\label{eq:individual}
\end{eqnarray}
where $\pi(\theta_i)$ is the prior distribution and $p(x_i | \theta_i)$ is the
likelihood function. If $N$ objects are analysed separately, inferences can
be made about the diversity of $\theta_i$ across the sample. However, these
inferences may be incorrect due to the implicit assumption that the prior for
all of the $\{\theta_i\}$ is independent \citep{2013arXiv1310.5177B}.
A hierarchical model may be used to overcome this problem.
Section~\ref{sec:single_object} describes the assumptions used for analysing
data from a single AGN.

In a hierarchical model, the prior for the parameters $\{\theta_i\}$ is created
by introducing hyperparameters $\alpha$ such that the joint prior for $\alpha$
and the $\{\theta_i\}$ is:
\begin{eqnarray}
p(\alpha, \theta_i) &=& p(\alpha)\prod_{i=1}^N p(\theta_i | \alpha)
\end{eqnarray}
The implied prior on the $\{\theta_i\}$ parameters is then
\begin{eqnarray}
p(\theta_i) &=& \int p(\alpha)\prod_{i=1}^N p(\theta_i | \alpha) \, d\alpha
\end{eqnarray}
which may imply that the $\theta$ values are ``clustered'' around some
typical value.

The posterior distribution for
$\btheta = \{\theta_1, ..., \theta_N\}$ and the hyperparameters $\alpha$
given the data 
$\bx = \{x_1, ..., x_N\}$, is then
\begin{eqnarray}
p(\alpha, \btheta | \bx) &\propto&
p(\alpha)p(\btheta|\alpha)p(\bx | \btheta, \alpha)\\
&=& p(\alpha)\prod_{i=1}^N f(\theta_i|\alpha)p(x_i | \theta_i)
\end{eqnarray}

It is possible to consider the data for all $N$ objects as one big data set
and infer the parameters of all objects simultaneously. However, this can be
computationally prohibitive. In some circumstances it is tractable to analyse
each object's data separately (using a common prior $\pi(\theta_i)$) and
then post-process the results, reconstructing what the results of the
hierarchical model would have been. In this letter we use this approach.

The marginal posterior distribution for the hyperparameters $\alpha$ is
\begin{eqnarray}
p(\alpha | \bx) &=&
\int p(\alpha, \btheta|\bx) \, d\btheta\\
&\propto& p(\alpha)\int \prod_{i=1}^N f(\theta_i|\alpha)p(x_i | \theta_i) d^N\theta\\
&\propto& p(\alpha) \prod_{i=1}^N \int f(\theta_i|\alpha)p(x_i | \theta_i) d\theta_i\\
&\propto& p(\alpha) \prod_{i=1}^N \int \frac{f(\theta_i|\alpha)}{\pi(\theta)}p(x_i | \theta_i) \pi(\theta_i)d\theta_i\\
&\propto& p(\alpha) \prod_{i=1}^N \mathds{E}\left[\frac{f(\theta_i|\alpha)}{\pi(\theta)}\right]\label{eq:combine}
\end{eqnarray}
where the expectation is taken with respect to the individual object posterior
of Equation~\ref{eq:individual}, and thus can be estimated using posterior
samples from the posterior distributions for the individual objects.
This result enables us to reconstruct the
posterior distribution for the hyperparameters even though the individual object
inferences were made without the hierarchical structure in the prior. This is
essentially an importance sampling approximation to the
full hierarchical model.

@@

\section{The Single Object Model}\label{sec:single_object}
If the continuum light curve of an AGN
is described by a function $y(t)$, then the line
light curve $l(t)$ is given by
\begin{eqnarray}
l(t) &=& A \int_\tau \Psi(\tau)\left[y(t - \tau) + C\right] \, d\tau
\end{eqnarray}
where $\Psi(\tau)$ is the transfer function (assumed to be normalised),
and $A$ and $C$ are response
coefficients. The idea of reverberation mapping is to use noisy measurements
of $y(t)$ and $l(t)$ to infer the transfer function $\Psi(\tau)$ or a summary
of it such as the mean lag $\bar{\tau} = \int \tau\Psi(\tau) \, d\tau$.
Throughout this letter we consider $y(t)$ and $l(t)$ in flux
units, as opposed to magnitudes, and consider $\bar{\tau}$ as
the definition of ``the lag'' of the AGN.

The posterior distribution for the lag $\bar{\tau}$ of a single object
$i$ may be obtained
by fitting the following model. We assume, for simplicity, that
the transfer function is uniform
between limits $a$ and $b$, where $b > a$:
\begin{eqnarray}
\Psi(\tau) &=& \left\{
\begin{array}{lr}
\frac{1}{b-a}, & \tau \in [a,b]\\
0, & \textnormal{otherwise}s
\end{array}\right.
\end{eqnarray}
and our goal is to measure the mean lag:
\begin{eqnarray}
\bar{\tau} &=& \int \tau \Psi(\tau) \, d\tau\\
&=& \frac{1}{2}(b-a)
\end{eqnarray}

Note that inferring $\bar{\tau}$ from the data requires that we
marginalise over an infinite number of nuisance parameters describing the
behaviour of $y(t)$ at unobserved times \citep{pancoast}.
The prior for the underlying time variation of the continuum emission is
a continuous autoregressive process of order 1, or a CAR(1) model. These models
have been studied extensively for AGN variability
\citep[e.g.][]{2009ApJ...698..895K, 2011ApJ...735...80Z, 2013ApJ...765..106Z}.
This marginalisation can be done
either analytically or inside MCMC. We used the latter approach for simplicity.
We discretised
time using ten time bins per day, and have a discrete continuum light curve
$\mathbf{y} = \{y_1, ..., y_n\}$, included in the model as a set of unknown
parameters. The prior for these parameters is:
\begin{eqnarray}
p(y_i | y_{i-1},  \mu, a, \beta) \sim \mathcal{N}
\left(\mu + a\left(y_{i-1} - \mu\right), \beta^2\right)
\end{eqnarray}
for $i \geq 2$. This is the discrete AR(1) model from time series theory.
$\mu$ describes the mean level of the continuum light curve, $\beta$ controls
the size of the short-term fluctuations, and $a$ controls the correlation
timescale (the timescale is given by $L = -1/\ln(a)$).

@@

To implement the MCMC for a single object,
we implemented our model in the STAN sampler \citep{nuts}
and Diffusive Nested Sampling \citep{dnest}. Due to the large number of
parameters, multiple modes, and strong correlations in the posterior distribution,
we found that Diffusive Nested Sampling was more effective than STAN.
Note that our single-object model is already an improvement over standard
cross-correlation techniques and is similar to the approach used by
\citep{2011ApJ...735...80Z}.

>>>>

% Bold symbols
\renewcommand{\btheta}{\{\theta_i\}}
\newcommand{\bx}{\{x_i\}}

@@

\section{Introduction}
Reverberation mapping (RM) is a key technique for the study
of active galactic nuclei (AGN). The technique is based on the temporal
fluctuations of the central continuum source, and the subsequent response
of the broad line region (BLR) emission. The time delay between the continuum
and the broad line fluctuations provides an estimate of the size of the BLR,
and can also be used to estimate the black hole mass \citep{peterson}.

RM is a observationally intensive, requiring observations of an AGN
over a period of
a few tens of days \citep{2013ApJ...769..128B}. As a result, many authors have
studied the data analysis techniques involved in RM, and substantial
advances have been made in recent years.
The data analysis methods introduced range from those that attempt to infer the
transfer function \citep[the distribution of lags in a single object,
e.g.][]{1995ApJ...440..166K, 2011ApJ...735...80Z},
the velocity-resolved transfer function
\citep{2010ApJ...720L..46B}, or the physical structure of the BLR itself
\citep{pancoast, arp151, pancoast2, 2013arXiv1310.3907L}.

There are many subtleties and challenges involved in reverberation mapping,
that we will ignore
for the purposes of this letter. We note two of them here for completeness. Firstly,
the mean lag $\bar{\tau} = \int \tau\Psi(\tau)\, d\tau$, where $\Psi(\tau)$
is the normalised transfer function, is not equal to $c$ times the mean radius of the BLR
matter distribution, $\bar{r} = \int \sqrt{x^2+y^2+z^2}\rho(x, y, z)\, d^3 \mathbf{x}$.
Secondly, the mean lag $\bar{\tau}$ is not equivalent to the peak of the
cross-correlation function, except in the case of very narrow
transfer functions. Direct physical modelling of the BLR resolves these
issues \citep{pancoast, arp151, 2013arXiv1310.3907L}.

@@

In this letter, we revisit the idea of ``stacked'' reverberation mapping (i.e.
using a sample of AGN with sparse data to infer something about the sample)
from a Bayesian perspective. We develop a Bayesian hierarchical model
\citep{loredo, kelly, extreme_deconvolution, 2012AJ....143...90S, 2012arXiv1208.3036L, 2013arXiv1310.5177B, 2013AJ....146....7B}.
which allows us to calculate the posterior distribution for hyperparameters
which describe the typical lag, and the diversity of the lags.

@@

\section{Combining Inferences About Multiple Objects}\label{sec:multiple}
Consider a sample of $N$ objects, each of which has parameters $\theta_i$.
If the $i$th object is analysed on its own, the inference about 
its parameters $\theta_i$ is described by a posterior distribution
\begin{eqnarray}
p(\theta_i | x_i) \propto \pi(\theta_i)p(x_i | \theta_i)\label{eq:individual}
\end{eqnarray}
where $\pi(\theta_i)$ is the prior distribution and $p(x_i | \theta_i)$ is the
likelihood function. If $N$ objects are analysed separately, inferences can
be made about the diversity of $\theta_i$ across the sample. However, these
inferences may be incorrect due to the implicit assumption that the prior for
all of the $\btheta$ is independent \citep{2013arXiv1310.5177B}.
A hierarchical model may be used to overcome this problem.

In a hierarchical model, the prior for the parameters $\btheta$ is created
by introducing hyperparameters $\alpha$ such that the joint prior for $\alpha$
and the $\btheta$ is:
\begin{eqnarray}
p(\alpha, \btheta) &=& p(\alpha)\prod_{i=1}^N p(\theta_i | \alpha)
\end{eqnarray}
The implied prior on the $\btheta$ parameters is then
\begin{eqnarray}
p(\btheta) &=& \int p(\alpha)\prod_{i=1}^N p(\theta_i | \alpha) \, d\alpha
\end{eqnarray}
which may imply that the $\theta$ values are ``clustered'' around some
typical value.

The posterior distribution for
$\btheta$ and the hyperparameters $\alpha$
given the data 
$\{x_1, ..., x_N\}$, is then
\begin{eqnarray}
p(\alpha, \btheta | \bx) &\propto&
p(\alpha)p(\btheta|\alpha)p(\bx | \btheta, \alpha)\\
&=& p(\alpha)\prod_{i=1}^N f(\theta_i|\alpha)p(x_i | \theta_i)
\end{eqnarray}

It is possible to consider the data for all $N$ objects as one big data set
and infer the hyperparameters $\alpha$ and the parameters of all objects $\btheta$
simultaneously. However, this can be
computationally prohibitive. In some circumstances it is tractable to analyse
each object's data separately (using a common prior $\pi(\theta_i)$) and
then post-process the results, reconstructing what the results of the
hierarchical model would have been. In this letter we use this latter approach.

The marginal posterior distribution for the hyperparameters $\alpha$ is
\begin{eqnarray}
p(\alpha | \bx) &=&
\int p(\alpha, \btheta|\bx) \, d\btheta\\
&\propto& p(\alpha)\int \prod_{i=1}^N f(\theta_i|\alpha)p(x_i | \theta_i) d^N\theta\\
&\propto& p(\alpha) \prod_{i=1}^N \int f(\theta_i|\alpha)p(x_i | \theta_i) d\theta_i\\
&\propto& p(\alpha) \prod_{i=1}^N \int \frac{f(\theta_i|\alpha)}{\pi(\theta)}p(x_i | \theta_i) \pi(\theta_i)d\theta_i\\
&\propto& p(\alpha) \prod_{i=1}^N \mathds{E}\left[\frac{f(\theta_i|\alpha)}{\pi(\theta)}\right]\label{eq:combine}
\end{eqnarray}
where the expectation is taken with respect to the individual object posterior
of Equation~\ref{eq:individual}, and thus can be estimated using posterior
samples from the posterior distributions for the individual objects.
This result enables us to reconstruct the
posterior distribution for the hyperparameters even though the individual object
inferences were made without the hierarchical structure in the prior. This is
essentially an importance sampling approximation to the
full hierarchical model. One drawback of this approach is that it implies an
independent (non-hierarchical) prior on any per-object nuisance parameters,
which can sometimes adversely affect the results \citep[e.g.][]{2013arXiv1310.5177B}.

@@

\section{The Single Object Model}\label{sec:single_object}
If the continuum light curve of an AGN
is described by a function $y(t)$, then the line
light curve $l(t)$ is given by
\begin{eqnarray}
l(t) &=& A \int_\tau \Psi(\tau)\left[y(t - \tau) + C\right] \, d\tau
\end{eqnarray}
where $\Psi(\tau)$ is the transfer function (assumed to be normalised),
and $A$ and $C$ are response
coefficients. The idea of reverberation mapping is to use noisy measurements
of $y(t)$ and $l(t)$ to infer the transfer function $\Psi(\tau)$ or a summary
of it such as the mean lag $\bar{\tau} = \int \tau\Psi(\tau) \, d\tau$.
Throughout this letter we consider $y(t)$ and $l(t)$ in flux
units, as opposed to magnitudes, and consider $\bar{\tau}$ as
the definition of ``the lag'' of an AGN.

The posterior distribution for the lag $\bar{\tau}$ of a single object
$i$ may be obtained
by fitting the following model. We assume, for simplicity, that
the transfer function is uniform
between limits $a$ and $b$, where $b > a$:
\begin{eqnarray}
\Psi(\tau) &=& \left\{
\begin{array}{lr}
\frac{1}{b-a}, & \tau \in [a,b]\\
0, & \textnormal{otherwise}
\end{array}\right.
\end{eqnarray}
and our goal is to measure the mean lag:
\begin{eqnarray}
\bar{\tau} &=& \int \tau \Psi(\tau) \, d\tau\\
&=& \frac{1}{2}(b-a).
\end{eqnarray}
Note that inferring $\bar{\tau}$ from the data requires that we
marginalise over an infinite number of nuisance parameters describing the
behaviour of $y(t)$ at unobserved times \citep{pancoast}.
The prior for the underlying time variation of the continuum emission is
a continuous autoregressive process of order 1, or a CAR(1) model. These models
have been studied extensively for AGN variability
\citep[e.g.][]{2009ApJ...698..895K, 2011ApJ...735...80Z, 2013ApJ...765..106Z}.
This marginalisation can be done
either analytically or inside MCMC. We used the latter approach for simplicity.
We discretised
time using ten time bins per day, and have a discrete continuum light curve
$\mathbf{y} = \{y_1, ..., y_n\}$, included in the model as a set of unknown
parameters. The prior for these parameters is:
\begin{eqnarray}
p(y_i | y_{i-1},  \mu, a, \beta) \sim \mathcal{N}
\left(\mu + a\left(y_{i-1} - \mu\right), \beta^2\right)
\end{eqnarray}
for $i \geq 2$. This is the discrete AR(1) model from time series theory.
$\mu$ describes the mean level of the continuum light curve, $\beta$ controls
the size of the short-term fluctuations, and $a$ controls the correlation
timescale (the timescale is given by $L = -1/\ln(a)$).

@@

To implement the MCMC for a single object,
we implemented our model in the STAN sampler \citep{nuts}
and Diffusive Nested Sampling \citep{dnest}. Due to the large number of
parameters, multiple modes, and strong correlations in the posterior distribution,
we found that Diffusive Nested Sampling was more effective than STAN.
Note that our single-object model is already an improvement over standard
cross-correlation techniques and is similar to the approach used by
\citet{2011ApJ...735...80Z}.
